# STANDARDIZING SCORES

**Chapter Links**

* [Chapter 4 Slide Show](http://tysonbarrett.com/EDUC-6600/Slides/u01_Ch4_Zscores.html#1)

* [Interactive Online App - Normal Curve](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_7_norm.html)

* [Interactive Online App - p Values from a Normal Curve](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_12_pvalue.html)


* [Interactive Online App - Statistical Significance](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_14_signif.html)



* [Cancer Dataset - SPSS format](https://usu.box.com/s/9c92zof5whb76bphmzxn3vqx5702qgq6)

**Assignment Links**

* [Unit 1 Assignment - Write Up Skeleton](https://usu.box.com/s/bn0m44cp5zrkev3ppjdilfsvic4bubvu)

* [Unit 1 Assignment - Rmd Skeleton](https://usu.box.com/s/zjjwexn827ziifxlcoe6emr2brfifk0x)

* [Inho's Dataset - Excel format](https://usu.box.com/s/hyky7eb24l6vvzj2xboedhcx1xolrpw1)


```{r global_options, include=FALSE}
# set global chunk options...  
#  this changes the defaults so you don't have to repeat yourself
knitr::opts_chunk$set(comment     = NA,
                      cache       = TRUE,
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE)
```



Required Packages 

```{r load_libraries}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(haven)        # Read in SPSS datasets
library(furniture)    # Nice tables (by our own Tyson Barrett)
```



Example: Cancer Experiment 

The `Cancer` dataset was introduced in [chapter 3][Example: Cancer Experiment].


```{r, include=FALSE}
cancer_raw <- haven::read_spss("data/cancer.sav")

cancer_clean <- cancer_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(trt = factor(trt,
                             labels = c("Placebo", 
                                        "Aloe Juice"))) %>% 
  dplyr::mutate(stage = factor(stage))
```






## Standardize Variables - Manually

You can manually create a stadradized version of the `age` variable.


First, you must find the **mean** and **standard deviation** of the `age` variable.

```{r}
cancer_clean %>%
  furniture::table1(age)
```


Second, write an equation to do the calculation.

```{r}
cancer_clean %>%
  dplyr::mutate(agez = (age - 59.6) / 12.9) %>% 
  dplyr::select(id, trt, age, agez)
```


## Standardize Variables - with the `scale()` funciton

A quicker way is to use a funciton.  Notice the differences due to rounding.

```{r}
cancer_new <- cancer_clean %>%
  dplyr::mutate(agez = (age - 59.6) / 12.9) %>% 
  dplyr::mutate(ageZ = scale(age))%>%
  dplyr::select(id, trt, age, agez, ageZ)

cancer_new
```

You can check that the new variable does in deed have mean of zero and spread of one.

```{r}
cancer_new %>% 
  furniture::table1(age, agez, ageZ,
                    digits = 8)
```


Both the mean and the standard deviation are different.

```{r}
cancer_new %>% 
  tidyr::gather(key = "variable",
                value = "value",
                age, ageZ) %>% 
  ggplot(aes(value)) +
  geom_histogram(bins = 8) +
  facet_grid(. ~ variable)
```

However, if you let the scale of the x-axis change, you see the shape of the two variables is identical.

```{r}
cancer_new %>% 
  tidyr::gather(key = "variable",
                value = "value",
                age, ageZ) %>% 
  ggplot(aes(value)) +
  geom_histogram(bins = 8) +
  facet_grid(. ~ variable, scale = "free_x")
```



