# t TEST FOR THE MEAN OF 1 SAMPLE

**Chapter Links**

* [Chapter 6 Slide Show](http://tysonbarrett.com/EDUC-6600/Slides/u02_Ch6_CI_tDist.html#1)

* [Interactive Online App - Confidence Intervals](http://digitalfirst.bfwpub.com/stats_applet/stats_applet_4_ci.html)

* [Cancer Dataset - SPSS format](https://usu.box.com/s/9c92zof5whb76bphmzxn3vqx5702qgq6)



**Assignment Links**

* [Unit 2 Assignment - Write Up Skeleton](https://usu.box.com/s/mr5ersj8oqu6mj3tyup697ljg2527p8h)

* [Unit 2 Assignment - Rmd Skeleton](https://usu.box.com/s/85s7t82tih6f06v8bpvjo0qt275gafsr)

* [Inho's Dataset - Excel format](https://usu.box.com/s/hyky7eb24l6vvzj2xboedhcx1xolrpw1)


```{r global_options, include=FALSE}
# set global chunk options...  
#  this changes the defaults so you don't have to repeat yourself
knitr::opts_chunk$set(comment     = NA,
                      cache       = TRUE,
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE)
```



Required Packages 

```{r load_libraries}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(haven)        # Read in SPSS datasets
```




Example: Cancer Experiment 

The `Cancer` dataset was introduced in [chapter 3][Example: Cancer Experiment].


```{r, include=FALSE}
cancer_raw <- haven::read_spss("data/cancer.sav")

cancer_clean <- cancer_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(trt = factor(trt,
                             labels = c("Placebo", 
                                        "Aloe Juice"))) %>% 
  dplyr::mutate(stage = factor(stage))
```





## 1 Sample Mean vs. historic control

example: Do the patients weigh more than 165 pounds at intake, on average?


```{r}
cancer_clean %>% 
  dplyr::pull(weighin) %>% 
  t.test(mu = 165)
```



## Change the Confidence Level

Find a 99% confience level for the population mean weight.

```{r}
cancer_clean %>% 
  dplyr::pull(weighin) %>% 
  t.test(mu = 165,
         conf.level = 0.99)
```


## Restrict to a Subsample

Do the patients with .dcoral[stage 3 & 4] cancer weigh more than 165 pounds at intake, on average?

```{r}
cancer_clean %>% 
  dplyr::filter(stage %in% c("3", "4")) %>% 
  dplyr::pull(weighin) %>% 
  t.test(mu = 165)
```


