# t TEST FOR THE DIFFERENCE IN 2 PAIRED MEANS

**Chapter Links**

* [Chapter 11 Slide Show (pdf)](http://tysonbarrett.com/EDUC-6600/Slides/u03_Ch11_paired.html#1)

* [Cancer Dataset - SPSS format](https://usu.box.com/s/9c92zof5whb76bphmzxn3vqx5702qgq6)



**Assignment Links**

* [Unit 3 Assignment - Write Up Skeleton .pdf](https://usu.box.com/s/vjcsotiqwu1mwnwgzbfyig6k451ymgow)

* [Unit 3 Assignment - R Directions .pdf](https://usu.box.com/s/ectr9zx8qfbbm59h0qcexjreje5r9aio)

* [Unit 3 Assignment - R Skeleton .Rmd](https://usu.box.com/s/k3vzw6nuq5tw66bxeptcyzth38pj69f9)

* [Inho's Dataset - Excel format](https://usu.box.com/s/hyky7eb24l6vvzj2xboedhcx1xolrpw1)




```{r global_options, include=FALSE}
# set global chunk options...  
#  this changes the defaults so you don't have to repeat yourself
knitr::opts_chunk$set(comment     = NA,
                      cache       = TRUE,
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE)
```



Required Packages 

```{r load_libraries}
library(tidyverse)    # Loads several very helpful 'tidy' packages
library(haven)        # Read in SPSS datasets
library(car)          # Companion for Applied Regression (and ANOVA)
```




Example: Cancer Experiment 

The `Cancer` dataset was introduced in [chapter 3][Example: Cancer Experiment].


```{r, include=FALSE}
cancer_raw <- haven::read_spss("data/cancer.sav")

cancer_clean <- cancer_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(trt = factor(trt,
                             labels = c("Placebo", 
                                        "Aloe Juice"))) %>% 
  dplyr::mutate(stage = factor(stage))
```



Check Means and SD’s

```{r}
cancer_clean %>% 
  dplyr::group_by(trt) %>% 
  furniture::table1(totalcin, totalcw4)
```


-------------------------------------------------------
-------------------------------------------------------


## Checking ASsumptions
